---
title: "Frequency-based feeds from observed service"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frequency-based feeds from observed service}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_gtfsrealtime <- requireNamespace("gtfsrealtime", quietly = TRUE)
```

Most of `gtfsrt2static` assembles **stop-times-based** feeds: one row per observed
stop event, preserving every run. `snapshot_frequencies()` does the opposite. It
collapses *many* observed runs of a route-direction into a compact,
**frequency-based** schedule — one representative trip per time window plus a
`frequencies.txt` headway — and emits one feed per reliability quantile:

- **structural** (p05): a free-flow lower bound,
- **median** (p50): a typical day,
- **reliable** (p95): a slow, worst-case upper bound.

The same quantile is applied to *both* travel time and headway, so the reliable
feed is slower and less frequent than the structural one. This is the shape you
want for accessibility and reliability analysis, where you compare routing under
optimistic, typical, and pessimistic service.

```{r setup}
library(gtfsrt2static)
library(data.table)
```

## The input: observed stop events

Every assembler in this package consumes the canonical *observed stop events*
table (see `?"observed-stop-events"`): one row per trip × stop × service date,
with absolute `POSIXct` arrival/departure times and a `provenance` label. You
get it from `snapshot_from_trip_updates()` (GTFS-Realtime Trip Updates) or
`snapshot_from_stop_times()` (gps2gtfs-derived stop times). For a first,
self-contained look we build a tiny one by hand: one route, six runs across a
morning peak, four stops, with per-run variation so the quantiles differ.

```{r synthetic-events}
base_date <- as.Date("2026-05-06") # a Wednesday
starts <- as.POSIXct(
  paste(base_date, c("07:00", "07:12", "07:25", "07:41", "08:02", "08:20")),
  tz = "UTC"
) # rising headways: 12, 13, 16, 21, 18 min
stops <- c("A1", "A2", "A3", "A4")
nominal <- c(0, 300, 720, 1080) # seconds from trip start
spread <- c(0, 40, -30, 90, 20, -20) # per-run travel-time noise

events <- rbindlist(lapply(seq_along(starts), function(i) {
  off <- nominal + c(0, 1, 2, 3) * spread[i] # later stops accumulate spread
  data.frame(
    trip_ref = paste0("run", i), route_ref = "A", shape_ref = NA_character_,
    direction_id = 0L, service_date = base_date,
    stop_ref = stops, stop_sequence = NA_integer_,
    arrival_time = starts[i] + off,
    departure_time = starts[i] + off + 20,
    provenance = "observed", vehicle_ref = paste0("veh", i),
    source = "gps", stringsAsFactors = FALSE
  )
}))
head(events)
```

## Assembling the feeds

`snapshot_frequencies()` needs the events, a **windows** definition (a
frequency feed is organised by time window), and — to be publishable — an
`agency` and a `stops` table with coordinates. `strict = TRUE` turns missing
agency/coordinate warnings into errors, acting as a publish gate.

```{r assemble}
stops_tbl <- data.frame(
  stop_id = stops,
  stop_lat = c(40.70, 40.71, 40.72, 40.73),
  stop_lon = c(-73.99, -73.98, -73.97, -73.96)
)

feeds <- snapshot_frequencies(
  events,
  windows = list(am_peak = c("07:00", "09:00")),
  stops = stops_tbl,
  agency = list(name = "Demo Transit", url = "https://example.org", timezone = "UTC"),
  strict = TRUE
)
names(feeds)
```

You get a named list of `gtfsio`-convention feed objects, one per quantile. Each
is a standard GTFS feed you write with `gtfsio::export_gtfs()`.

The `frequencies.txt` carries the observed headway for the window, per quantile:

```{r frequencies}
feeds$structural$frequencies
feeds$reliable$frequencies
```

The representative `stop_times` are **offsets from a `00:00:00` trip start**
(the `exact_times = 0` convention: only relative offsets matter), clamped
non-decreasing so arrivals never go backwards along the trip:

```{r stop-times}
feeds$median$stop_times
```

The three feeds order as you'd expect — the reliable trip takes longer to reach
its last stop, and runs at a longer headway, than the structural one:

```{r ordering}
clock_secs <- function(x) {
  p <- tstrsplit(x, ":", fixed = TRUE)
  as.integer(p[[1]]) * 3600L + as.integer(p[[2]]) * 60L + as.integer(p[[3]])
}
last_arrival <- sapply(feeds, function(f) {
  max(clock_secs(f$stop_times$arrival_time))
})
headway <- sapply(feeds, function(f) f$frequencies$headway_secs)
data.frame(scenario = names(feeds), last_arrival_secs = last_arrival, headway_secs = headway)
```

Each feed also carries a programmatic publish gate — `snapshot_publishable()`
reads the `publishable` / `publish_blockers` attributes stamped on it:

```{r publishable}
snapshot_publishable(feeds$median)$publishable
```

## From a real GTFS-Realtime feed

In practice the events come from a real feed. Here is the exact same pipeline on
a bundled snapshot of New York City MTA Trip Updates (via the `gtfsrealtime`
package) — no manual preprocessing beyond selecting a few routes. The reader
emits warnings about duplicate trip descriptors in this particular snapshot;
those concern the source feed's identifiers, not the conversion, so we suppress
them here.

```{r real-events, eval = has_gtfsrealtime}
tu <- suppressWarnings(gtfsrealtime::read_gtfsrt_trip_updates(
  system.file("nyc-trip-updates.pb.bz2", package = "gtfsrealtime"),
  timezone = "America/New_York"
))
tu <- as.data.table(tu)
tu <- tu[route_id %in% c("M1", "M3", "M4")] # a small Manhattan slice

events_nyc <- snapshot_from_trip_updates(tu, tz = "America/New_York")
table(events_nyc$provenance)
```

A single poll is *prediction-dominated* — most stops have not been passed yet,
so they are labelled `predicted-last` rather than `observed`; an archive
collected over time (see `rt_collect()`) shifts that balance. The frequency
assembly treats both as timed passages.

Assembling frequency feeds is identical to the synthetic case. Supplying an
`agency` clears the agency blocker; the only remaining blocker is stop
coordinates, which live in the operator's *static* GTFS `stops.txt`:

```{r real-assemble, eval = has_gtfsrealtime}
feeds_nyc <- snapshot_frequencies(
  events_nyc,
  windows = list(pm_peak = c("16:00", "19:00"), evening = c("19:00", "22:00")),
  agency = list(name = "MTA New York City Transit",
                url = "https://www.mta.info", timezone = "America/New_York"),
  strict = FALSE # no coordinates yet -> a non-publishable analysis feed
)
snapshot_publishable(feeds_nyc$median)$publishable
snapshot_publishable(feeds_nyc$median)$blockers
```

To make it publishable, pass a `stops` data frame (`stop_id` / `stop_lat` /
`stop_lon`) read from the static feed, exactly as in the synthetic example:

```{r real-publish, eval = FALSE}
stops <- gtfsio::import_gtfs("path/to/mta_static.zip")$stops
feeds_nyc <- snapshot_frequencies(
  events_nyc,
  windows = list(pm_peak = c("16:00", "19:00"), evening = c("19:00", "22:00")),
  stops = stops[, c("stop_id", "stop_lat", "stop_lon")],
  agency = list(name = "MTA New York City Transit",
                url = "https://www.mta.info", timezone = "America/New_York"),
  strict = TRUE
)
```

Run on a coordinate-complete 25-route Manhattan slice (537 trips), all three
feeds pass the MobilityData `gtfs-validator` (v6.0.0) with **zero ERROR
notices**; the residual notices are the WARNINGs expected for a historical,
prediction-heavy single snapshot (fast-travel data-quality signals, an expired
calendar). The resulting feeds are ordinary GTFS zips that `gtfstools`,
`tidytransit`, and any other GTFS tool read directly.
