---
title: "Frequency-based feeds from observed service"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frequency-based feeds from observed service}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_gtfsrealtime <- requireNamespace("gtfsrealtime", quietly = TRUE)
```

Most of `gtfsrt2static` assembles **stop-times-based** feeds: one row per observed
stop event, preserving every run. `rt2s_frequencies()` does the opposite. It
collapses *many* observed runs of a route-direction into a compact,
**frequency-based** schedule — one representative trip per time window plus a
`frequencies.txt` headway — and emits one feed per reliability quantile:

- **structural** (p05): a free-flow lower bound,
- **median** (p50): a typical day,
- **reliable** (p95): a slow, worst-case upper bound.

By default the same quantile is applied to *both* travel time and headway, so the
reliable feed is slower and less frequent than the structural one. This is the
shape you want for accessibility and reliability analysis, where you compare
routing under optimistic, typical, and pessimistic service.

That default is a convenience, not a constraint. Passing a *list* instead of a
numeric vector decouples the two sides, which matters as soon as a scenario is
not simply "everything optimistic": a free-flow running time at a *typical*
frequency is a common estimand, and coupling would drag a p05 headway along with
the p05 travel time.

```r
quantiles = list(
  structural = c(travel = 0.05, headway = 0.50),  # free-flow, typical frequency
  median     = 0.50,                              # still coupled - shorthand
  reliable   = 0.95
)
```

An omitted side inherits the one you gave, so `c(headway = 0.5)` is legal for a
scenario whose travel times do not come from an observed quantile at all — see
the anchored section at the end.

```{r setup}
library(gtfsrt2static)
library(data.table)
```

## The input: observed stop events

Every assembler in this package consumes the canonical *observed stop events*
table (see `?"observed-stop-events"`): one row per trip × stop × service date,
with absolute `POSIXct` arrival/departure times and a `provenance` label. You
get it from `rt2s_events_from_trip_updates()` (GTFS-Realtime Trip Updates) or
`rt2s_events_from_stop_times()` (gps2gtfs-derived stop times). For a first,
self-contained look we build a tiny one by hand: one route, six runs across a
morning peak, four stops, with per-run variation so the quantiles differ.

```{r synthetic-events}
base_date <- as.Date("2026-05-06") # a Wednesday
starts <- as.POSIXct(
  paste(base_date, c("07:00", "07:12", "07:25", "07:41", "08:02", "08:20")),
  tz = "UTC"
) # rising headways: 12, 13, 16, 21, 18 min
stops <- c("A1", "A2", "A3", "A4")
nominal <- c(0, 300, 720, 1080) # seconds from trip start
spread <- c(0, 40, -30, 90, 20, -20) # per-run travel-time noise

events <- rbindlist(lapply(seq_along(starts), function(i) {
  off <- nominal + c(0, 1, 2, 3) * spread[i] # later stops accumulate spread
  data.frame(
    trip_ref = paste0("run", i), route_ref = "A", shape_ref = NA_character_,
    direction_id = 0L, service_date = base_date,
    stop_ref = stops, stop_sequence = NA_integer_,
    arrival_time = starts[i] + off,
    departure_time = starts[i] + off + 20,
    provenance = "observed", vehicle_ref = paste0("veh", i),
    source = "gps", stringsAsFactors = FALSE
  )
}))
head(events)
```

## Assembling the feeds

`rt2s_frequencies()` needs the events, a **windows** definition (a
frequency feed is organised by time window), and — to be publishable — an
`agency` and a `stops` table with coordinates. `strict = TRUE` turns missing
agency/coordinate warnings into errors, acting as a publish gate.

```{r assemble}
stops_tbl <- data.frame(
  stop_id = stops,
  stop_lat = c(40.70, 40.71, 40.72, 40.73),
  stop_lon = c(-73.99, -73.98, -73.97, -73.96)
)

feeds <- rt2s_frequencies(
  events,
  windows = list(am_peak = c("07:00", "09:00")),
  stops = stops_tbl,
  agency = list(name = "Demo Transit", url = "https://example.org", timezone = "UTC"),
  strict = TRUE
)
names(feeds)
```

You get a named list of `gtfsio`-convention feed objects, one per quantile. Each
is a standard GTFS feed you write with `gtfsio::export_gtfs()`.

The `frequencies.txt` carries the observed headway for the window, per quantile:

```{r frequencies}
feeds$structural$frequencies
feeds$reliable$frequencies
```

The representative `stop_times` are **offsets from a `00:00:00` trip start**
(the `exact_times = 0` convention: only relative offsets matter), clamped
non-decreasing so arrivals never go backwards along the trip:

```{r stop-times}
feeds$median$stop_times
```

The three feeds order as you'd expect — the reliable trip takes longer to reach
its last stop, and runs at a longer headway, than the structural one:

```{r ordering}
clock_secs <- function(x) {
  p <- tstrsplit(x, ":", fixed = TRUE)
  as.integer(p[[1]]) * 3600L + as.integer(p[[2]]) * 60L + as.integer(p[[3]])
}
last_arrival <- sapply(feeds, function(f) {
  max(clock_secs(f$stop_times$arrival_time))
})
headway <- sapply(feeds, function(f) f$frequencies$headway_secs)
data.frame(scenario = names(feeds), last_arrival_secs = last_arrival, headway_secs = headway)
```

Each feed also carries a programmatic publish gate — `rt2s_publishable()`
reads the `publishable` / `publish_blockers` attributes stamped on it:

```{r publishable}
rt2s_publishable(feeds$median)$publishable
```

## What the feeds were built from

The feeds tell you what was written; they do not tell you what was *considered*
and left out. `rt2s_resolved_grid()` does — one row per candidate
`(route_ref, direction_id, window)` and scenario:

```{r grid}
grid <- rt2s_resolved_grid(feeds)
grid
```

`ratio` is the running-time ratio applied (`NA` here, because an observed
pattern carries its own travel-time quantile and no ratio exists);
`headway_secs` is the headway actually written to `frequencies.txt` and
`headway_source` says whether it came from the observations or from a
`headways=` override; `emitted` is whether the trip reached that scenario's
`trips.txt`; and `drop_reason` explains the cells that did not.

This exists so a pipeline can reconcile its own cell accounting against the
feed. The grid is built from the candidate set captured **before** any drop
stage runs, so its row count does not change as cells fall out — which is what
makes a stage-by-stage funnel check possible:

```{r grid-funnel}
# every candidate cell is accounted for, in every scenario
table(grid$scenario, grid$drop_reason, useNA = "ifany")

# and the emitted rows match what was actually written -- these feeds were built
# without extra_trips=, so the grid accounts for every row of trips.txt
identical(
  sort(grid[scenario == "median" & emitted == TRUE]$trip_id),
  sort(feeds$median$trips$trip_id)
)
```

That last identity holds **because no `extra_trips` were passed**. The grid has
one row per candidate `(route, direction, window)` cell, and individually-timed
trips are not cells; if you add them, reconcile against
`c(grid[emitted == TRUE]$trip_id, <the ids you supplied>)` instead. The next
section shows that.

A cell is dropped for one of two reasons. `"no_stop_pattern"` means the group
had a headway but no served (or, when anchoring, no baseline) stop pattern.
`"no_ratio"` means `scaling_missing = "drop"` removed it — from *every*
scenario, so the feeds keep one shared trip set (see the anchored section
below).

One case is deliberately **not** a drop: a cell can be `emitted` with
`headway_secs = NA`. The representative trip and its `stop_times` were written,
but no `frequencies.txt` row was, because that group's headway quantile was
missing or non-positive. Such a row carries no `drop_reason`, so do not infer
"emitted" from "has a headway".

## Mixing in individually-timed trips

Real service is not purely frequency-based. Some cells cannot honestly be
written as one representative trip plus a repeating headway — an hour with three
irregular departures, a peak extra, a cell whose evidence is a handful of exact
passages rather than a rhythm. GTFS handles this directly: **only trips listed in
`frequencies.txt` are frequency-based**, and every other trip is read from
`stop_times` as exact scheduled times. A feed may contain both.

`extra_trips=` is how you supply the second kind. It takes a named list keyed by
scenario — the same names as `quantiles`, so a misspelled scenario is an error
rather than a silent no-op — each element a `list(trips=, stop_times=)` whose
times are **absolute clock strings**, not offsets from `00:00:00`:

```{r extra-trips}
# the inverse of clock_secs() above: seconds since midnight -> "HH:MM:SS"
secs_clock <- function(s) {
  sprintf("%02d:%02d:%02d", s %/% 3600, (s %% 3600) %/% 60, s %% 60)
}

extra <- list(
  trips = data.frame(
    route_id = "A", trip_id = c("A_exact_1", "A_exact_2"), direction_id = 0L
  ),
  stop_times = rbindlist(lapply(c("A_exact_1", "A_exact_2"), function(id) {
    start <- if (id == "A_exact_1") 8 * 3600 + 2100 else 8 * 3600 + 3300
    off <- c(0, 300, 720, 1080)
    data.frame(
      trip_id = id,
      arrival_time = secs_clock(start + off),
      departure_time = secs_clock(start + off + 20),
      stop_id = stops,
      stop_sequence = seq_along(stops)
    )
  }))
)

mixed <- rt2s_frequencies(
  events,
  windows = list(am_peak = c("07:00", "09:00")),
  stops = stops_tbl,
  agency = list(name = "Demo Transit", url = "https://example.org", timezone = "UTC"),
  strict = TRUE,
  # the two late departures are only defensible for the median scenario here
  extra_trips = list(median = extra)
)
mixed$median$trips
```

The extra trips are in `trips.txt` and `stop_times.txt`, and deliberately **not**
in `frequencies.txt` — that absence is what makes them exact-time trips:

```{r extra-frequencies}
mixed$median$frequencies
```

Three properties worth knowing:

- **Their stops and routes are added to `stops.txt` and `routes.txt`**, but every
  `stop_id` must already be known (from `stops`, or from an emitted pattern) and
  every `route_id` must be an emitted or baseline route. A dangling reference is
  an invalid feed, so it is rejected rather than filled in with a placeholder —
  the lenient missing-coordinates path exists for observed data with genuinely
  unknown positions, not for a typo in a table you supplied.
- **No cross-scenario invariant is imposed.** A scenario may supply more, fewer
  or no extra trips than another. Exact-time evidence legitimately differs by
  scenario — a scheduled scenario draws it from the timetable while the others
  draw it from observed passages — so requiring matching id sets would reject
  correct data. The shared-trip-set guarantee stays scoped to the *generated*
  frequency trips, where `scaling_missing` enforces it.
- **They are not rows of `rt2s_resolved_grid()`.** The grid is one row per candidate
  cell; extra trips are not cells, and adding them would produce rows whose
  `route_ref`, `window`, `ratio`, `headway_secs` and `drop_reason` are all `NA`.
  Since you supplied the trips, you already own their ids, so the funnel still
  closes:

```{r extra-funnel}
mixed_grid <- rt2s_resolved_grid(mixed)
identical(
  sort(c(
    mixed_grid[scenario == "median" & emitted == TRUE]$trip_id,
    extra$trips$trip_id
  )),
  sort(mixed$median$trips$trip_id)
)
```

## From a real GTFS-Realtime feed

In practice the events come from a real feed. Here is the exact same pipeline on
a bundled snapshot of New York City MTA Trip Updates (via the `gtfsrealtime`
package) — no manual preprocessing beyond selecting a few routes. The reader
emits warnings about duplicate trip descriptors in this particular snapshot;
those concern the source feed's identifiers, not the conversion, so we suppress
them here.

```{r real-events, eval = has_gtfsrealtime}
tu <- suppressWarnings(gtfsrealtime::read_gtfsrt_trip_updates(
  system.file("nyc-trip-updates.pb.bz2", package = "gtfsrealtime"),
  timezone = "America/New_York"
))
tu <- as.data.table(tu)
tu <- tu[route_id %in% c("M1", "M3", "M4")] # a small Manhattan slice

events_nyc <- rt2s_events_from_trip_updates(tu, tz = "America/New_York")
table(events_nyc$provenance)
```

A single poll is *prediction-dominated* — most stops have not been passed yet,
so they are labelled `predicted-last` rather than `observed`; an archive
collected over time (see `rt2s_collect()`) shifts that balance. The frequency
assembly treats both as timed passages.

Assembling frequency feeds is identical to the synthetic case. Supplying an
`agency` clears the agency blocker; the only remaining blocker is stop
coordinates, which live in the operator's *static* GTFS `stops.txt`:

```{r real-assemble, eval = has_gtfsrealtime}
feeds_nyc <- rt2s_frequencies(
  events_nyc,
  windows = list(pm_peak = c("16:00", "19:00"), evening = c("19:00", "22:00")),
  agency = list(name = "MTA New York City Transit",
                url = "https://www.mta.info", timezone = "America/New_York"),
  strict = FALSE # no coordinates yet -> a non-publishable analysis feed
)
rt2s_publishable(feeds_nyc$median)$publishable
rt2s_publishable(feeds_nyc$median)$blockers
```

To make it publishable, pass a `stops` data frame (`stop_id` / `stop_lat` /
`stop_lon`) read from the static feed, exactly as in the synthetic example:

```{r real-publish, eval = FALSE}
stops <- gtfsio::import_gtfs("path/to/mta_static.zip")$stops
feeds_nyc <- rt2s_frequencies(
  events_nyc,
  windows = list(pm_peak = c("16:00", "19:00"), evening = c("19:00", "22:00")),
  stops = stops[, c("stop_id", "stop_lat", "stop_lon")],
  agency = list(name = "MTA New York City Transit",
                url = "https://www.mta.info", timezone = "America/New_York"),
  strict = TRUE
)
```

Run on a coordinate-complete 25-route Manhattan slice (537 trips), all three
feeds pass the MobilityData `gtfs-validator` (v6.0.0) with **zero ERROR
notices**; the residual notices are the WARNINGs expected for a historical,
prediction-heavy single snapshot (fast-travel data-quality signals, an expired
calendar). The resulting feeds are ordinary GTFS zips that `gtfstools`,
`tidytransit`, and any other GTFS tool read directly.

## Anchoring on a planned static feed

Everything above **reconstructs** the stop pattern from the observations. That is
right when no usable published pattern exists. But it has a consequence worth
being explicit about: because the stops and their order are derived from what was
observed, two feeds built from different observations can differ in their
*networks*, not just their service levels. If you then compare accessibility
between a scheduled and an observed feed, "service got slower" is confounded with
"the network changed".

When the operator does publish a usable feed, anchor on it instead. Set
`pattern_source = "baseline"` and the stop pattern comes from the published
`stop_times` — one canonical (modal) pattern per route-direction, chosen by
`rt2s_baseline_patterns()` — with only two numbers varying per route and window: a
**running-time ratio** and a **headway**. Every scenario then emits the identical
trip set over the identical stops.

```{r anchored, eval = FALSE}
static <- gtfsio::import_gtfs("path/to/operator_static.zip")
windows <- list(am_peak = c("06:00", "09:00"), interpeak = c("09:00", "16:00"))

# Running-time ratios you estimated yourself: observed duration relative to
# planned, per route, direction, window and scenario. The package does not
# estimate these - the definition of "duration", the coverage gates and any
# plausibility bounds belong with whoever measured them.
ratios <- data.frame(
  route_ref = "B62", direction_id = 0L, window = "am_peak",
  scenario = c("scheduled", "structural", "median", "reliable"),
  ratio = c(1.00, 0.87, 1.05, 1.31)
)

# The "scheduled" scenario's frequency comes from the timetable itself, not from
# observations, so take it off the planned feed and inject it as an override.
sched <- rt2s_baseline_headways(static, windows = windows)
sched$scenario <- "scheduled"

feeds <- rt2s_frequencies(
  events,
  windows = windows,
  quantiles = list(
    scheduled  = c(headway = 0.50),                 # travel side unused here
    structural = c(travel = 0.05, headway = 0.50),
    median     = c(travel = 0.50, headway = 0.50),
    reliable   = c(travel = 0.95, headway = 0.95)
  ),
  baseline = static,
  pattern_source = "baseline",
  scaling = ratios,
  headways = sched
)
```

Three things to know about this mode:

- **`events` is still required.** Only the pattern source changes; headways, the
  calendar's weekdays and the feed's date range still come from the observations.
- **The travel side of `quantiles` is inert.** The pattern is the published one
  scaled by `ratio`, so `rt2s_obs_travel_times()` is never called. `quantiles` still
  defines *which* scenarios exist — it stays the single source of scenario
  identity, and `scaling`/`headways` may only refer to its names.
- **The trip set is shared by construction.** Ratios are resolved for the whole
  (trip × scenario) grid before any feed is built, so a cell missing a ratio
  leaves *every* scenario rather than just the one that lacks it. That is what
  `scaling_missing` controls, and why its default is to error rather than quietly
  produce feeds you cannot compare. Under `scaling_missing = "drop"` the removed
  cells stay visible in `rt2s_resolved_grid()` with `drop_reason = "no_ratio"`, so a
  drop is something you can audit rather than something you have to notice.

The identity contract is the thing to get right: `events$route_ref` must carry the
same identifier as the baseline's `trips.route_id` (or `routes.route_short_name`,
via `route_key`), and `direction_id` must agree. A completely disjoint key set is
an error naming examples from both sides.

One incidental advantage: a route that visits the same stop twice (a loop, or a
branch that doubles back) is handled correctly here, because the pattern is a
literal published trip. The reconstructing path collapses repeated visits into one
position and warns about it.
