---
title: "Get started with gtfsrt2static"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with gtfsrt2static}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_deps <- requireNamespace("gtfsrealtime", quietly = TRUE) &&
  requireNamespace("gtfstools", quietly = TRUE)
knitr::opts_chunk$set(eval = has_deps)
```

`gtfsrt2static` turns observations of transit service (GTFS-Realtime Trip
Updates, or stop times inferred from GPS by
[`gps2gtfs`](https://e-kotov.github.io/gps2gtfs/)) into a static GTFS feed of
the service that actually ran. This article takes one bundled snapshot of New
York City bus Trip Updates through the whole path:

1. read the protobuf with [`gtfsrealtime`](https://cran.r-project.org/package=gtfsrealtime);
2. reduce it to *observed stop events*, the table every assembler consumes;
3. check the events;
4. scaffold a GTFS feed from them;
5. write the feed and read it back with
   [`gtfstools`](https://ipea.github.io/gtfstools/).

```{r setup, eval = TRUE}
library(gtfsrt2static)
library(data.table)
```

## Read the Trip Updates

`gtfsrealtime` parses GTFS-Realtime protobuf (single snapshots, compressed
files, or a daily ZIP of archived polls) into one row per stop-time update.
Pass the timezone the agency operates in. The reader warns about duplicate trip
descriptors in this particular snapshot; those concern the source feed's
identifiers, not the conversion, so we suppress them.

```{r read}
tu <- suppressWarnings(gtfsrealtime::read_gtfsrt_trip_updates(
  system.file("nyc-trip-updates.pb.bz2", package = "gtfsrealtime"),
  timezone = "America/New_York"
))
tu <- as.data.table(tu)
dim(tu)

tu <- tu[route_id == "M15"] # one Manhattan bus route keeps the example small
tu[1:3, .(trip_id, route_id, stop_sequence, stop_id, arrival_time)]
```

## Reduce to observed stop events

`rt2s_events_from_trip_updates()` keeps, for each trip, service date and stop
visit, the latest report, and labels it `observed` when it was issued after the
vehicle left the stop and `predicted-last` otherwise. Skipped stops and
canceled trips become explicit rows. Both input paths of the package,
Trip Updates and gps2gtfs stop times, end in this same table; see
`?"observed-stop-events"` for its columns.

```{r events}
events <- rt2s_events_from_trip_updates(tu, tz = "America/New_York")
events[1:3, .(trip_ref, route_ref, service_date, stop_ref, stop_sequence,
              arrival_time, provenance)]
table(events$provenance)
```

A single poll is *prediction-dominated*: most stops have not been passed yet,
so they are labelled `predicted-last` rather than `observed`. An archive
collected over the day with `rt2s_collect()` shifts that balance towards
`observed`.

## Check the events

`rt2s_events_validate()` checks that a table conforms to the observed stop
events schema and returns it, so it fits in a pipe; an error names the first
violation. The converters and assemblers already call it, so run it yourself on
events you build or modify by hand.

```{r validate}
events <- rt2s_events_validate(events)
```

## Scaffold a feed

With no planned feed to inherit from, `rt2s_scaffold()` synthesizes every
spec-required GTFS file, with linked identifiers. Two things cannot come from
Trip Updates: agency metadata, which you supply, and stop coordinates, which
live in the operator's static `stops.txt`.

```{r scaffold}
feed <- rt2s_scaffold(
  events,
  agency = list(
    name = "MTA New York City Transit",
    url = "https://www.mta.info",
    timezone = "America/New_York"
  )
)
names(feed)
head(feed$stop_times)
```

The times are now GTFS clock strings on the service day's clock. Each feed
carries its own publish gate. This one is not publishable yet, and says why:

```{r publishable}
rt2s_publishable(feed)
```

Pass a `stops` table (`stop_id`, `stop_lat`, `stop_lon`) taken from the static
feed, and set `strict = TRUE` to turn any remaining gap into an error:

```{r strict, eval = FALSE}
static_stops <- gtfsio::import_gtfs("path/to/mta_static.zip")$stops
feed <- rt2s_scaffold(
  events,
  agency = list(name = "MTA New York City Transit",
                url = "https://www.mta.info", timezone = "America/New_York"),
  stops = static_stops[, c("stop_id", "stop_lat", "stop_lon")],
  strict = TRUE
)
```

When you do have the operator's planned feed, use
`rt2s_assemble(events, baseline = ...)` instead. The realized feed then keeps the
official route, trip, stop and service identifiers, so planned and realized
service join directly.

## Write it and read it back

The feed follows the [`gtfsio`](https://r-transit.github.io/gtfsio/)
convention, so `gtfsio::export_gtfs()` writes a standard GTFS zip, and any GTFS
tool reads it. Here `gtfstools` computes each trip's duration from the file:

```{r export}
path <- tempfile(fileext = ".zip")
gtfsio::export_gtfs(feed, path)

gtfs <- gtfstools::read_gtfs(path)
head(gtfstools::get_trip_duration(gtfs, unit = "min"))
```

A trip that appears for a few minutes covers only the stops still ahead of the
bus when the snapshot was taken.

## Where next

- [From raw GPS to a realized GTFS feed](https://e-kotov.github.io/gtfsrt2static/articles/pipeline.html): the same
  assembly fed by `gps2gtfs` from vehicle trajectories, in both scaffold and
  baseline mode, with a planned-versus-realized comparison.
- `vignette("frequency-feeds")`: collapsing many observed runs into
  frequency-based feeds at three reliability quantiles.
