---
title: "From raw GPS to a realized GTFS feed"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
pkgs <- c("gtfsrealtime", "gps2gtfs", "gtfsio", "gtfstools")
has_deps <- all(vapply(pkgs, requireNamespace, logical(1), quietly = TRUE))
knitr::opts_chunk$set(eval = has_deps)
```

This article follows transit observations from raw vehicle positions to a
static GTFS feed of the service that actually ran, across four packages:

| Step | Package | Functions |
|---|---|---|
| Parse GTFS-Realtime protobuf | [`gtfsrealtime`](https://cran.r-project.org/package=gtfsrealtime) | `read_gtfsrt_positions()` |
| Infer trips and stop visits from coordinates | [`gps2gtfs`](https://e-kotov.github.io/gps2gtfs/) | `g2g_stops_from_gtfs()`, `g2g_terminals_from_gtfs()`, `g2g_extract_trips_and_stop_times()`, `g2g_shapes_from_trips()` |
| Assemble a compliant feed | `gtfsrt2static` | `rt2s_events_from_stop_times()`, `rt2s_assemble()` |
| Write, read and analyse GTFS | [`gtfsio`](https://r-transit.github.io/gtfsio/), [`gtfstools`](https://ipea.github.io/gtfstools/) | `export_gtfs()`, `read_gtfs()`, ... |

The simple example scaffolds a feed from one day of a bus's GPS pings with no
planned feed. The advanced example matches GTFS-Realtime positions to an
operator's planned feed and compares planned with realized times.

## Setup

`gps2gtfs` is not on CRAN. Install it, and `gtfsrt2static`, from GitHub:

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak(c("e-kotov/gps2gtfs", "e-kotov/gtfsrt2static", "gtfsrealtime", "gtfstools"))
```

```{r setup}
library(gps2gtfs)
library(gtfsrt2static)
library(data.table)
```

## Vehicle Positions are GPS pings

GTFS-Realtime Vehicle Positions report where each vehicle is. `gtfsrealtime`
parses them, here from one bundled snapshot of the New York City bus feed:

```{r read-vp}
vp <- gtfsrealtime::read_gtfsrt_positions(
  system.file("nyc-vehicle-positions.pb.bz2", package = "gtfsrealtime"),
  timezone = "America/New_York"
)
vp <- as.data.table(vp)
vp[1:3, .(vehicle_id, trip_id, route_id, latitude, longitude, timestamp, speed)]
```

These are the column names `gps2gtfs` expects, so the reader's output goes into
the extractor with no adapter. Columns it does not use, such as `trip_id`,
`route_id` and `stop_id`, are passed through. This feed leaves `speed` empty.
`gps2gtfs` detects dwell from pings with `speed == 0`, so with such a feed it
warns and treats every ping as moving.

One snapshot holds one ping per vehicle:

```{r one-ping}
nrow(vp)
uniqueN(vp$vehicle_id)
```

A trajectory needs many polls of the same vehicle. Archive the endpoint over a
day with `rt2s_collect()`, then read the daily ZIP with the same function. The
examples below use trajectories that ship with `gps2gtfs` instead.

## Simple example: scaffold a feed from GPS pings

`gps2gtfs` bundles one day of GPS pings from a bus on route 654 between Kandy
and Digana, Sri Lanka, with the route's stops and terminals. There is no
planned GTFS feed for it.

### Extract trips, stop visits and shapes

```{r g2g-extract}
result <- g2g_extract_trips_and_stop_times(
  gps_data = g2g_data_gps,
  terminals_data = g2g_data_terminals,
  stops_data = g2g_data_stops,
  terminals_buffer_radius = 50,
  stops_buffer_radius = 30,
  stops_extended_buffer_radius = 50,
  stop_direction_map = c("Kandy-Digana" = "BT01", "Digana-Kandy" = "BT02"),
  return_trajectory = TRUE,
  diagnostics_warn = FALSE
)
result$trips[, .(trip_id, start_terminal, end_terminal, start_time, end_time)]

shapes <- g2g_shapes_from_trips(result$trajectory)
```

The [gps2gtfs Get started](https://e-kotov.github.io/gps2gtfs/articles/gps2gtfs.html)
article explains these arguments and the coverage diagnostics.

### Convert to observed stop events

`rt2s_events_from_stop_times()` turns the stop visits into the *observed stop
events* table that every `gtfsrt2static` assembler consumes. `gps2gtfs` models
one route per run, so the route identity is stamped on. `shape_ref_prefix`
must match the `shape_id_prefix` of `g2g_shapes_from_trips()` (`"SHP_"` by
default) so each trip links to its driven shape.

```{r events-gps}
events <- rt2s_events_from_stop_times(
  result$stop_times,
  route_ref = "654",
  source = "gps",
  shape_ref_prefix = "SHP_"
)
events[1:3, .(trip_ref, route_ref, shape_ref, service_date, stop_ref,
              stop_sequence, arrival_time, provenance)]
```

Each trip has no official identifier, so it gets a synthetic one, unique
across service dates.

### Scaffold the feed

With no planned feed, `rt2s_assemble()` scaffolds every required file. Agency
metadata and stop coordinates must be supplied, and `strict = TRUE` makes any
gap an error, so a feed that returns is free of known spec-required gaps. The
bundled pings are stored in UTC, so UTC is the feed's timezone here; with real
data, give the agency's own.

```{r scaffold}
stops <- data.frame(
  stop_id = as.character(g2g_data_stops$stop_id),
  stop_name = g2g_data_stops$address,
  stop_lat = g2g_data_stops$latitude,
  stop_lon = g2g_data_stops$longitude
)

feed <- rt2s_assemble(
  events,
  agency = list(name = "Route 654 demo", url = "https://example.org",
                timezone = "UTC"),
  stops = stops,
  shapes = shapes,
  strict = TRUE
)
names(feed)
feed$trips[]
rt2s_publishable(feed)$publishable
```

### Write it and analyse it with gtfstools

```{r scaffold-export}
path <- tempfile(fileext = ".zip")
gtfsio::export_gtfs(feed, path)

gtfs <- gtfstools::read_gtfs(path)
merge(
  gtfstools::get_trip_duration(gtfs, unit = "min"),
  gtfstools::get_trip_speed(gtfs, unit = "km/h")[, .(trip_id, speed)],
  by = "trip_id"
)
```

`get_trip_speed()` measures each trip along its shape, which here is the road
the bus actually drove. The same zip opens in
[`tidytransit`](https://r-transit.github.io/tidytransit/) and other GTFS tools.

## Advanced example: realized service against a planned feed

When the operator publishes a planned feed, and the Vehicle Positions carry its
`trip_id`, the realized feed can keep the official identifiers. It then joins
directly to the planned feed. The example uses a small two-stop route, so that
every step fits on screen.

### The planned feed

A planned feed with route `R1`, terminals `A` and `B`, two intermediate
stops, and two trips: `CS_1` from A to B at 08:00, and `CS_2` back at 08:30.

```{r planned}
lat <- seq(7.290, 7.320, length.out = 4)
lon <- seq(80.630, 80.660, length.out = 4)
planned_times <- c("08:00:00", "08:04:00", "08:08:00", "08:12:00",
                   "08:30:00", "08:34:00", "08:38:00", "08:42:00")

planned <- gtfsio::new_gtfs(list(
  agency = data.table(agency_id = "DT", agency_name = "Demo Transit",
                      agency_url = "https://example.org",
                      agency_timezone = "UTC"),
  routes = data.table(route_id = "R1", agency_id = "DT",
                      route_short_name = "1", route_type = 3L),
  calendar = data.table(service_id = "daily",
                        monday = 1L, tuesday = 1L, wednesday = 1L,
                        thursday = 1L, friday = 1L, saturday = 1L,
                        sunday = 1L,
                        start_date = 20260601L, end_date = 20260630L),
  stops = data.table(stop_id = c("A", "S1", "S2", "B"),
                     stop_name = c("Terminal A", "Stop 1", "Stop 2",
                                   "Terminal B"),
                     stop_lat = lat, stop_lon = lon),
  trips = data.table(route_id = "R1", service_id = "daily",
                     trip_id = c("CS_1", "CS_2"), direction_id = 0:1),
  stop_times = data.table(
    trip_id = rep(c("CS_1", "CS_2"), each = 4),
    arrival_time = planned_times,
    departure_time = planned_times,
    stop_id = c("A", "S1", "S2", "B", "B", "S2", "S1", "A"),
    stop_sequence = rep(1:4, 2)
  )
))

planned_zip <- tempfile(fileext = ".zip")
gtfsio::export_gtfs(planned, planned_zip)
```

### Positions annotated with the planned trip

These are the parsed Vehicle Positions of the bus on 6 June 2026, polled every
five minutes: one run from A to B, a layover ping without a trip, and the run
back. The `trip_id` column is what `read_gtfsrt_positions()` returns when the
feed annotates positions with the planned trip.

```{r positions}
t0 <- as.POSIXct("2026-06-06 08:00:00", tz = "UTC")
positions <- data.table(
  vehicle_id = "7482",
  latitude = c(lat, 7.321, rev(lat)),
  longitude = c(lon, 80.661, rev(lon)),
  timestamp = t0 + c(0, 300, 600, 900, 1200, 1800, 2100, 2400, 2700),
  speed = c(0, 20, 20, 0, 0, 0, 20, 20, 0),
  trip_id = c(rep("CS_1", 4), NA, rep("CS_2", 4))
)
```

### Stops and terminals from the planned feed

`gps2gtfs` derives both tables from the planned feed. The stops table holds
the route's non-terminal stops, labelled by the terminal their direction
starts from, so no direction map is needed:

```{r stops-terminals}
terminals <- g2g_terminals_from_gtfs(planned, route_id = "R1")
stops <- g2g_stops_from_gtfs(planned, route_id = "R1")
terminals
stops
```

### Extract along the planned trips

`trip_col = "trip_id"` takes the trips from the annotation instead of inferring
them from terminal crossings (the *fast path*). The annotation is kept as
`provided_trip_id`.

```{r extract-fast}
result <- g2g_extract_trips_and_stop_times(
  gps_data = positions,
  terminals_data = terminals,
  stops_data = stops,
  terminals_buffer_radius = 100,
  stops_buffer_radius = 100,
  stops_extended_buffer_radius = 150,
  trip_col = "trip_id",
  diagnostics_warn = FALSE
)
result$stop_times[, .(trip_id, provided_trip_id, stop_id, arrival_time)]
```

### Assemble against the baseline

`rt2s_events_from_stop_times()` uses a `provided_trip_id` column as the trip
identity when one is present. Given the planned feed as `baseline`,
`rt2s_assemble()` keeps its agency, routes and stops, matches the observed
trips by `trip_id`, and replaces `stop_times.txt` with the observed times for
the service day:

```{r baseline}
events <- rt2s_events_from_stop_times(result$stop_times)
unique(events$trip_ref)

realized <- rt2s_assemble(events, baseline = planned_zip)
realized$trips
realized$stop_times
```

### Compare planned and realized

Both feeds share trip and stop identifiers, so `gtfstools` reads both and a
join gives the delay at each stop. Join on `trip_id` and `stop_id`, not on
`stop_sequence`: for `gps2gtfs` input the realized `stop_sequence` counts the
visits in observed order, while the planned one numbers every stop of the
pattern, including the terminals. (In a loop that serves a stop twice, the
stop alone does not identify the visit.)

```{r compare}
realized_zip <- tempfile(fileext = ".zip")
gtfsio::export_gtfs(realized, realized_zip)

secs <- function(path) {
  g <- gtfstools::read_gtfs(path)
  gtfstools::convert_time_to_seconds(g, file = "stop_times")$stop_times
}
comparison <- merge(
  secs(planned_zip)[, .(trip_id, stop_id, planned = arrival_time_secs)],
  secs(realized_zip)[, .(trip_id, stop_id, realized = arrival_time_secs)],
  by = c("trip_id", "stop_id")
)
comparison[, delay_min := (realized - planned) / 60]
comparison
```

Each run was one minute late at its first intermediate stop and two at its
second.

## Where next

- `vignette("frequency-feeds", package = "gtfsrt2static")`: collapse many
  observed runs into frequency-based feeds at three reliability quantiles, and
  anchor them on a planned feed.
- [GTFS-Realtime Vehicle Positions to GTFS](https://e-kotov.github.io/gps2gtfs/articles/gtfs-realtime.html):
  the other ways `gps2gtfs` gets stops and terminals, from route geometry or
  from the positions themselves.
- `?"observed-stop-events"`: the table every input path converges on.
